Machine Learning Methods in Functional Programming and Interactive Theorem Proving

### 2015-09-24 TransferReport/report.tex
Added transfer report

% Based on the LLNCS.DEM demonstration file of
% the LaTeX macro package from Springer-Verlag
% for Lecture Notes in Computer Science,
% version 2.4 for LaTeX2e as of 16. April 2010
%
\documentclass{llncs}
%
\usepackage{makeidx}  % allows for indexgeneration
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{enumitem}
%
\begin{document}
%
\pagestyle{headings}  % switches on printing of running heads

\mainmatter              % start of the contributions
%
\title{Scaling Automated Theory Exploration}
%
\author{Chris Warburton \and Ekaterina Komendantskaya}
%
\institute{University of Dundee,\\
\texttt{http://tocai.computing.dundee.ac.uk}}

@@

\begin{abstract}
We investigate the \textbf{theory exploration} (TE)
paradigm for computer-assisted Mathematics and identify limitations and
improvements for current approaches. Unlike the theorem-proving paradigm,
which requires user-provided conjectures, TE performs an open-ended
search for theorems satisfying given criteria. We see promise in TE for
identifying new abstractions and connections in libraries of software
and proofs, but realising this potential requires more scalable
algorithms than presently used.
\end{abstract}
%
\section{Introduction}

@@

>>>>

\documentclass[]{article}

\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{enumitem}
\usepackage{paralist}
\usepackage{csquotes}
\usepackage[affil-it]{authblk}

\begin{document}

\pagestyle{headings}  % switches on printing of running heads

\title{Machine Learning Methods in Functional Programming and Interactive Theorem Proving}

\author{Chris Warburton}

\affil{University of Dundee,\\
\texttt{http://tocai.computing.dundee.ac.uk}}

@@

\begin{abstract}
Since their inception, computers have been applied to automate and assist work in mathematics, with established fields including numerical computation, computer algebra, (online) communication and so on. Here we focus on their use in two areas: formal proof and statistics; and how new approaches are bringing these seemingly disparate disciplines closer together.
\end{abstract}

\section{Mechanised Proof}

When computers are used to aid the development of formal proofs, it is usually within either the \emph{interactive theorem proving} (ITP) or \emph{automated theorem proving} (ATP) framework. Such theorem proving approaches are characterised as being \emph{goal-driven}: particular conjectures must be provided by the user before attempts are made to (dis)prove them. In addition to these traditional approaches, we also consider the recently proposed \emph{theory exploration} (TE) paradigm, which does not use (explicit) goals, and of which ATP and ITP are crucial components.

\subsection{Interactive Theorem Proving}

ITP systems are built around an automated \emph{proof checker}. This is a relatively simple task to mechanise \cite{boyer1994qed}, although care must be taken during implementation since bugs may cause inconsistencies to arise in the resulting logic TODO: Reference some example, eg. proof of false in Coq. A popular approach to mitigating this risk, as well as making meta-level analysis of the system more tractable, is known as the \emph{de Bruijn criterion}: implementing a very simple ``core'' checker, and having all other features be translated into a form checkable by this core \cite[\S~3]{barendregt2005challenge} TODO Maybe Chapter 18. Proof-Assistants Using Dependent Type Systems (Henk Barendregt, Herman Geuvers). instead?. This is the approach taken by, for example, the Isabelle and Coq systems \cite{nipkow2002isabelle} \cite{bertot2013interactive}.

In the ITP setting, proving a conjecture is equivalent to producing an input to the checker which \begin{inparaenum}[a)]
  \item is accepted and
  \item has that conjecture as its conclusion.
\end{inparaenum}

TODO: How closely does Isabelle match the type-theoretic approach of Coq, Agda, etc.? Can we make this less ``fluffy''?

Of course, \emph{producing} such proofs is much more difficult than checking them; as the name suggests, ITP leaves this task up to the user. To facilitate easier proving, many ITP systems such as Coq and Agda use a pure functional programming language as their proof format. Besides theoretical elegance, due to the \emph{Curry-Howard correspondence} \cite{wadler2015propositions}, there are practical benefits to turning theorem proving into programming. Firstly, many techniques developed in software engineering such as meta-programming (usually in the form of \emph{tactics}), encapsulation and modularity can be directly applied to the proving process. Secondly, the process of \emph{software verification}, a common task for formal proof, can be achieved in a very straightforward way: the software can be implemented inside the proof language, certified correct by the proof checker, then \emph{extracted} into an executable form.

Despite these conveniences, ITP still requires enormous effort for non-trivial proof or verification tasks (e.g. see \cite{hales2015formal}). The most obvious way to mitigate this problem is by implementing powerful tactics to automate as much of a proof as possible. The most striking example of this approach is the \emph{sledgehammer} system in Isabelle/HOL \cite{journals/iandc/MengQP06}, which translates goals into a form suitable for a multitude of ATP systems and invokes them all in parallel to see if any can provide a proof which Isabelle's core checker will accept.

\subsection{Automated Theorem Proving}

ATP is the autonomous search for proofs in some logic. Every ATP system is built around a complete or partial decision procedure such as \emph{resolution} \cite[\S~9.6]{Russell:2003:AIM:773294}, usually augmented by heuristics and often adjustable using parameters. Most standalone ATP systems use a first-order logic \cite[\S~10]{Russell:2003:AIM:773294}, since many tasks in higher-order logics are undecidable (e.g. \cite{huet1973undecidability}).

ATP is a classic field of research in artificial intelligence, dating back to the founding of the field at the 1956 Dartmouth conference. Even by that time Newell and Simon has developed their Logic Theory Machine \cite{newell1956logic}, which was subsequently able to prove theorems like those in Principia Mathematica \cite{newell1958elements}.

In a similar way to ITP, the central problem of ATP is simple to state but unfeasible to execute: search through the space of provable statements until the given conjecture is found; the path from conjecture to the initial axioms consitutes the proof. However, this search space grows exponentially in the length of the proofs, which is unfortunate since proof length has been proposed as an approximate measure of how interesting a theorem is \cite[\S~10.2.1]{colton2012automated}.

\subsection{Theory Exploration}

Our main subject of investigation is the use of computers for augmenting and automating the process of \emph{theory exploration} (TE). This is the discovery of ``interesting'' patterns, relationships and structure in a formal system.

Buchberger identifies TE as \textquote{building up (large) mathematical knowledge bases in an efficient, reliable, well-structured, re-usable, and flexible way} \cite{buchberger2004algorithm}, and considers it \textquote{the natural paradigm for using automated or computer-supported proving systems} \cite{buchberger2000theory}. In other words, the ITP and ATP approaches described above are fundamental components of any TE system, but the primary focus is on \emph{generating} conjectures.

The most striking difference between TE and theorem proving is the lack of an explicit goal (e.g. a user-provided conjecture to prove). Instead, the \emph{implicit} goal is more open-ended: the discovery of ``interesting'' conclusions from the given premises. Of course, a key question to ask is what do we mean by ``interesting''? There is no single answer, although many approximate measures have been proposed. The choice of what counts as ``interesting'' has been used to classify various TE implementations \cite{warburtonscaling}, along with their approaches to term generation, well-formedness and proof.

In the case of Buchberger's \texttt{Theorema}, the first self-confessed TE system, only the questions of well-formedness and proof were tackled; specifically by the use of types and ATP algorithms, respectively. Similar to the approach of ITP, the more difficult tasks (term generation and the determination of their interest) is left up to the user.

\iffalse

  task of theory exploration, and its existing software implementations, by Buchberger as complementary to  takes place after concrete definitions have been established (eg. by the process of \emph{theory formation})  identified by . Theory exploration is similar to \emph{experimental mathematics} process of identifying \emph{interesting} consequences of a given theory.

 - Relation to Science
  - Testable/falsifiable hypotheses are like evaluable terms (or, more generally, conjectures which can be decided, using a reasonable amount of resources).
 - Relation to AI tasks: exploring surroundings, etc.

Theory exploration must take place after \emph{theory formation}

Major fields include computer algebra and theorem proving.

- Statistics is another area that's less straightforward than normal numerical computing, since there is subjectivity and judgement involved in the answering of questions.

Computer algebra systems, such as Mathematica, are
There arithmetic, has a Initially used for ``computation'', ie. evaluation of expressions,
especially in arithmetic numerical The application to evaluation

Theory formation: Alison? Others.
Theory exploration: Buchberger, Moa in Isabelle, Koen in Haskell. Others?
Theorem proving: Well-trodden: first-order ATP, higher-order ITP, functional programming
Communication: Latex, Wikis, APIs, communicating with aliens

\subsection{Exploration in Artificial Intelligence and Machine Learning}

The theme of goals and exploration which occur in theorem proving and TE have analogues in the domains of artificial intelligence (AI) and machine learning (ML). In  in an commonly appearing in the form of the \emph{exploration vs exploitation problem} in goal-driven systems.

Machine Learning

Symbolic AI: ATP, theory exploration?

Integrating statistical and symbolic

Machine learning over structured data:
1D is common: parsing natural language
2D is common; images
Trees are fractal
Backpropagation through structure
LSTM with recursive structure
Most work tries to identify structure; we already have it

One way to apply fixed-size machine learning algorithms to arbitrarily-sized recursive structures is to use a \emph{distributed representation}. These mix information from all parts of a structure together into a fixed number of bits, storing an \emph{approximation} whose accuracy depends on the size of the value and the amount of storage used.

CURRENT PROGRESS

Divide and conquor for theory exploration
Does it help?

Clustering and feature extraction

FUTURE WORK

QuickSpec: extend or extinguish?

Improve and find other use cases/scenarios for clustering and feature extraction

Other directions for Theory Exploration?

How about systems based on term rewriting, logic programming, etc.?

\begin{abstract}
We investigate the \textbf{theory exploration} (TE)
paradigm for computer-assisted Mathematics and identify limitations and
improvements for current approaches. Unlike the theorem-proving paradigm,
which requires user-provided conjectures, TE performs an open-ended
search for theorems satisfying given criteria. We see promise in TE for
identifying new abstractions and connections in libraries of software
and proofs, but realising this potential requires more scalable
algorithms than presently used.
\end{abstract}

@@

\fi
